---
title: "EDS 212"
subtitle: "Day 1: Course intro & basic math review"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
    css: xaringan-themer.css
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#003660",
  header_font_google = google_font("Sanchez"),
  text_font_google   = google_font("Nunito Sans", "300", "300i"),
  code_font_google   = google_font("Roboto Mono"),
  base_font_size = "18px",
  text_font_size = "26px",
  table_border_color = "#ffffff",
  table_row_border_color = "#ffffff",
  table_row_even_background_color = "ffffff",
  text_slide_number_font_size = "0.7em",
)
```

## Welcome to EDS 212 - Essential Math in Environmental Data Science

- **Teacher:** Allison Horst (ahorst@ucsb.edu)
- **Teaching assistant:** 
- **Course hours:** 8am - 5pm PST, Monday - Friday
- **Location:** NCEAS MEDS Classroom

---

## Course description

- **Units:** 2
- **Grading:** Pass/No Pass 
- **Description:** Quantitative skills and understanding are critical when working with, understanding, analyzing and gleaning insights from environmental data. In the intensive EDS 212 course, students will refresh fundamental skills in basic math (algebra, uni- and multivariate functions, units and unit conversions), derivative and integral calculus, differential equations, linear algebra, and reading, writing and evaluating logical operations.

---

## Daily schedule
|                   |                                                            |
|-------------------|------------------------------------------------------------|
| **Time (PST)**    | **Activity**                                               |
| 8:00am - 9:00am   | *Lecture 1 (60 min)*                                       |
| 9:00am - 9:15am   | *Break 1 (15 min)*                                         |
| 9:15am - 10:15am  | *Interactive Session 1 (60 min)*                           |
| 10:15am - 11:00am | *Flex time (45 min)*                                       |
| 11:00am - 12:15pm | *Break 2 (75 min)*                                         |
| 12:15pm - 1:15pm  | *Lecture 2 (60 min)*                                       |
| 1:15pm - 2:15pm   | *Interactive Session 2 (60 min)*                           |
| 2:15pm - 3:15pm   | *Flex time (60 min)*                                       |
| 3:15pm - 5:00pm   | *Interactive Session 3: Group & challenge tasks (105 min)* |

---

## Topics overview

### Day 1: Course introduction & math basics refresher
### Day 2: Derivatives
### Day 3: Integrals
### Day 4: Differential equations
### Day 5: Linear and Boolean algebra

---

class: center, middle

## Why am I taking a math class? 

(or, *Why do I have to know math when a computer will do math for me?*)

---

## A quantitative brain warm-up

### Diverse work & academic histories

- Focus on applications to environmental data science
- Specifically relevant for MEDS projects, coursework
- Start building computational skills & get used to some tools

### Even short gaps = stale brains

- Refresh quantitative thinking skills
- Refresh math essentials like units, unit conversions, gut checks, orders of magnitude, notation, language, and more

---

## Essential math in environmental data science



---

# Math Expressions

You can write LaTeX math expressions inside a pair of dollar signs, e.g. &#36;\alpha+\beta$ renders $\alpha+\beta$. You can use the display style with double dollar signs:

```
$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$
```

$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$

Limitations:

1. The source code of a LaTeX math expression must be in one line, unless it is inside a pair of double dollar signs, in which case the starting `$$` must appear in the very beginning of a line, followed immediately by a non-space character, and the ending `$$` must be at the end of a line, led by a non-space character;

1. There should not be spaces after the opening `$` or before the closing `$`.

1. Math does not work on the title slide (see [#61](https://github.com/yihui/xaringan/issues/61) for a workaround).

---

# R Code

```{r comment='#'}
# a boring regression
fit = lm(dist ~ 1 + speed, data = cars)
coef(summary(fit))
```

---

# R Plots

```{r cars, fig.height=4, dev='svg'}
par(mar = c(4, 4, 1, .1))
plot(cars, pch = 19, col = 'darkgray', las = 1)
abline(fit, lwd = 2)
```

---

# Tables

If you want to generate a table, make sure it is in the HTML format (instead of Markdown or other formats), e.g.,

```{r}
knitr::kable(head(iris), format = 'html')
```

---


---

# Some Tips

- There are several ways to build incremental slides. See [this presentation](https://slides.yihui.org/xaringan/incremental.html) for examples.

- The option `highlightLines: true` of `nature` will highlight code lines that start with `*`, or are wrapped in `{{ }}`, or have trailing comments `#<<`;

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          highlightLines: true
    ```

    See examples on the next page.

---

# Some Tips


.pull-left[
An example using a leading `*`:

    ```r
    if (TRUE) {
    ** message("Very important!")
    }
    ```
Output:
```r
if (TRUE) {
* message("Very important!")
}
```

This is invalid R code, so it is a plain fenced code block that is not executed.
]

.pull-right[
An example using `{{}}`:

````
`r ''````{r tidy=FALSE}
if (TRUE) {
*{{ message("Very important!") }}
}
```
````
Output:
```{r tidy=FALSE}
if (TRUE) {
{{ message("Very important!") }}
}
```

It is valid R code so you can run it. Note that `{{}}` can wrap an R expression of multiple lines.
]

---

# Some Tips

An example of using the trailing comment `#<<` to highlight lines:

````markdown
`r ''````{r tidy=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```
````

Output:

```{r tidy=FALSE, eval=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```

---

# Some Tips

When you enable line-highlighting, you can also use the chunk option `highlight.output` to highlight specific lines of the text output from a code chunk. For example, `highlight.output = TRUE` means highlighting all lines, and `highlight.output = c(1, 3)` means highlighting the first and third line.

````md
`r ''````{r, highlight.output=c(1, 3)}
head(iris)
```
````

```{r, highlight.output=c(1, 3), echo=FALSE}
head(iris)
```

Question: what does `highlight.output = c(TRUE, FALSE)` mean? (Hint: think about R's recycling of vectors)

---
