---
title: "EDS 212"
subtitle: "Day 3 Part 2 - Linear algebra fundamentals"  
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

## Linear algebra introduction

Linear algebra "branch of mathematics concerning linear equations" (Wikipedia), sometimes also described as the math of vectors & matrices. 

It is a fundamental part of data science (and how computers understand & process data), and useful for describing environmental processes.

As [Dr. Jason Brownlee](https://machinelearningmastery.com/gentle-introduction-linear-algebra/) writes, "Linear algebra is the mathematics of data."

---

## The building blocks of linear algebra

- **scalar:** a value without direction, representing magnitude

- **vector:** an ordered list of values, representing magnitude *and* direction (physics) or values for an observation or variable (data science)

- **matrix:** a table of values

---

## Applications of linear algebra in environmental sciences

- Dimensional reduction
- Population matrix models
- Optimization

## Applications of linear algebra in data science

- Array programming / vectorized code
- Machine learning
- For loops

---

## Let's start with vectors

### Where are vectors in EDS?

Everywhere. 

Vectors are lists of values used to describe different features or variables of interest. For example, if you are trying to model *fish size* based on length (cm) and mass (g), then for a fish with length 32 cm weighing 281 g, you might describe that by: 

$$\vec{F} = [32, 281]$$

---

From Wickham's *Advanced R* chapter on [Data Structures](http://adv-r.had.co.nz/Data-structures.html): "Under the hood, a data frame is a list of equal-length vectors."

```{r}
food <- c("banana", "apple", "carrot")
meal <- c("breakfast", "snack", "lunch")
food_mass_g <- c(14.8, 19.2, 11.5)

squirrel_meals <- data.frame(food, meal, food_mass_g) 
squirrel_meals # Returns the data frame
```

---

## Notation

Vectors are indicated with an arrow over the vector name:

$\vec{A} = [1,5]$

$\vec{AB} = [10.2, 3.1]$

---

## Vector addition & subtraction

Just add or subtract the corresponding pieces.

If: $\vec{A} = [2, 6]$ and $\vec{B} = [11, 10]$, then: 

$$\vec{A} - \vec{B} = [2-11, 6-10] = [-9, -4]$$

What does this look like graphically? **Let's draw it!**

---

## Scalar multipliers

You can multiply any vector by a scalar (constant). This will not change the *direction* of the vector - it will only change the *magnitude* of the vector. 

Example: $\vec{A} = [2,4]$

$$\vec{B} = 3\vec{A} = [3*2, 3*4] = [6, 12]$$

What does this look like graphically? **Let's draw it!**

---

## In R: 

We create vectors in R using the `c()` function (for combine or concatenate), and can perform operations on numeric vectors using basic operators.

```{r}
# Create vectors A and B: 
A <- c(1,2)
B <- c(5,9)
```

---

## Vector addition and subtraction in R

Just do it! For $\vec{A}=[1,2]$ and $\vec{B}=[5,9]$:

```{r}
# Addition: 
A + B

# Subtraction: 
A - B
```
---

## Scalar multiplication in R

Just do it!

For $\vec{A}=[1,2]$, calculate $4\vec{A}$:

```{r}
4*A
```
---

## Vectors of > 3 elements

$\vec{M}=c(2,4,1,8,6)$

Is as valid as describing a "point" in multivariate space as a vector with two "coordinates" -- it's just difficult for us to visualize and conceptualize since our brain only happily deals with 3 dimensions. 

---

## Dot product

For vectors $\vec a$ and $\vec b$, their dot product is:

$$\vec a \cdot \vec b = \sum a_i b_i$$
**In words:** The dot product is the sum of elements of each vector multiplied together, and is a measure of how close the vectors "point" in the same direction

---

## Dot product example

For $\vec a=[2,-1,0]$ and $\vec b=[9,3,-4]$:

$$\vec a \cdot \vec b = (2)(9)+(-1)(3)+(0)(-4) =15$$

This becomes very useful when describing systems of equations (tomorrow).

---

## What happens when we have orthogonal vectors? 

Sketch a quick graph, then find the dot product, of the following vector combinations: 

1. $\vec a=[0,4]$ and $\vec b =[6,0]$

2. $\vec c=[-3,1]$ and $\vec d=[2,6]$

What is the value of the dot product for orthogonal vectors? 

---

## More on vector fundamentals:

Optional: watch 3Brown1Blue's great 10 min recording on [Vectors (Ch 1 Essense of Linear Algebra)](https://www.youtube.com/watch?v=fNk_zzaMoSs&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=1).

### Vector addition & scalar multiplication are the basis of most linear algebra! 

