---
title: "Day 4 Tasks & Activities"
subtitle: 'Matrices and matrix algebra, exploring and summarizing data'
output: 
  html_document:
    css: activities_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(GGally)
```

## 1. A Leslie matrix example (small groups)

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are given the following:

- Each breeding adult will produce, on average, 600 eggs
- 25% of eggs will survive to become fry
- 10% of fry will survive to become juveniles
- 40% of juveniles will survive to become adults
- Adult survival rate year to year is 80%

1. With your group, draw the Leslie matrix model for this population structure.

2. With your group, given an initial population (Year 0) that has 0 eggs, 40,000 fry, 600 juveniles, and 450 adults, **by hand** project the population structure (i.e. the count in each life stage) in Year 1.

## 2. Fork & clone a repo to check your Leslie matrix, and make projections

- Fork and clone [this repo](https://github.com/allisonhorst/eds212-leslie-matrix), which contains the Leslie matrix for the example above. 

- Explore the code and outputs. Does the Year 1 population projection align with what you calculated? 

- Add code to the .Rmd to expand the projections to years 4, 5, and 6 (it current projects through year 3)

- Save, commit, pull, then push your changes.

```{r, include = FALSE}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.25, 0, 0, 0, 0, 0.10, 0, 0, 0, 0, 0.4, 0.8), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out:
fish_leslie

# Initial population: 
fish_year0 <- c(0, 40000, 600, 450)

# Population structure in year 1:
fish_year1 <- fish_leslie %*% fish_year0
fish_year1

# Population structure in year 2: 
fish_year2 <- fish_leslie %*% fish_year1
fish_year2

# Population structure in year 3: 
fish_year3 <- fish_leslie %*% fish_year2
fish_year3
```


## 3. Make a repo, do some stuff

- Create a new repo on GitHub named `eds212-day4-activities`
- Clone to create a version-controlled R Project

### A. Do some data exploration & summarizing in R

- Create a new R Markdown document in your project, save as `r-exploring``
- Attach the `tidyverse`, `skimr`, and `GGally` packages in the setup code chunk
- Run `View(diamonds)` in the Console to look at the built-in `diamonds` dataset in R
- Explore the dataset using the functions `names()`, `dim()`, `summary()`, `head()`, `tail()`, and `skimr::skim()`
- Create a pairs plot using `ggpairs()`
- Create a basic `ggplot` scatterplot of diamond price (`price`) as a function of size (`carat`)
- Create a histogram showing the distribution of values in the `carat` column (recall: a histogram only requires a single variable, e.g. `aes(x = carat)`)
- Save, stage, commit, pull, then push your changes back to your repo

### B. Now do that in Python

- Open Anaconda Navigator and launch Jupyter Notebooks
- Navigate to your project folder (`eds212-day4-activities`)
- Create a new Jupyter notebook (Python3), rename `py-exploring`
- In your notebook, in a new code chunk attach the seaborn and pandas packages:
    - `import seaborn as sns`
    - `import pandas as pd`
- Import the diamonds dataset in seaborn, storing as diamonds:
    - `diamonds = sns.load_dataset('diamonds')`
- Explore the dataset using the following (replacing 'df_name' with the name of your data frame):
    - `df_name.columns`
    - `df_name.head()`
    - `df_name.tail()`
    - `df_name.describe()`
    - `df_name.size`
    - `sns.pairplot(df_name)`
- Create a pairplot using `sns.pairplot(df_name)`
- Create a histogram of all values in the `carat` column using `sns.histplot(data=df_name, x="variable_name")`
- Save your .ipynb
- Back in RStudio, you should see your saved .ipynb (check the time modified)
- Stage, commit, pull, then push all changes to your GitHub repo
- Check to see that your changes were stored remotely

## End



