---
title: "Day 4 Tasks & Activities"
subtitle: 'Matrices and matrix algebra, exploring and summarizing data'
output: 
  html_document:
    css: activities_theme.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(GGally)
```

## 1. A Leslie matrix example (small groups)

A population of fish we're interested in has four life stages: eggs (E), fry (F), juvenile (J), breeding adult (A). You are given the following:

- Each breeding adult will produce, on average, 600 eggs
- 25% of eggs will survive to become fry
- 10% of fry will survive to become juveniles
- 40% of juveniles will survive to become adults
- Adult survival rate year to year is 80%

1. With your group, draw the Leslie matrix model for this population structure.

2. With your group, given an initial population (Year 0) that has 0 eggs, 40,000 fry, 600 juveniles, and 450 adults, **by hand** project the population structure (i.e. the count in each life stage) in Year 1.

## 2. Fork & clone a repo to check your Leslie matrix, and make projections

- Fork and clone [this repo](https://github.com/allisonhorst/eds212-leslie-matrix), which contains the Leslie matrix for the example above. 

    - Go to the link above. Press Fork in upper right, then Create Fork. 
    - Once in your fork, press the green 'Code' button
    - Copy the URL to your clipboard
    - Back in RStudio, create a new project (with version control)
    - Where prompted, paste the URL & choose where you'll save your directory 
    - Create project
    - Open the .Rmd to explore & run Leslie Matrix code

- Explore the code and outputs. Does the Year 1 population projection align with what you calculated? 

- Add code to the .Rmd to expand the projections to years 4, 5, and 6 (it current projects through year 3)

- **Using the command line:** Save, stage, commit, then push your changes.

```{r, include = FALSE}
# Make the Leslie Matrix:
fish_leslie <- matrix(c(0, 0, 0, 600, 0.25, 0, 0, 0, 0, 0.10, 0, 0, 0, 0, 0.4, 0.8), nrow = 4, ncol = 4, byrow = TRUE)

# Check it out:
fish_leslie

# Initial population: 
fish_year0 <- c(0, 40000, 600, 450)

# Population structure in year 1:
fish_year1 <- fish_leslie %*% fish_year0
fish_year1

# Population structure in year 2: 
fish_year2 <- fish_leslie %*% fish_year1
fish_year2

# Population structure in year 3: 
fish_year3 <- fish_leslie %*% fish_year2
fish_year3
```


## 3. Data exploration & summarizing in R

- Create a new Project
- Add a new Quarto document in your project, save as `r-exploring`
- Set up your local and remote git repo by running `usethis::use_git()` and then `usethis::use_github()`
- Attach the `tidyverse`, `skimr`, and `GGally` packages in the setup code chunk
- Run `View(diamonds)` in the Console to look at the built-in `diamonds` dataset in R
- Explore the dataset using the functions `names()`, `dim()`, `summary()`, `head()`, `tail()`, and `skimr::skim()`
- Create a pairs plot using `ggpairs()`
- Create a basic `ggplot` scatterplot of diamond price (`price`) as a function of size (`carat`)
- Create a histogram showing the distribution of values in the `carat` column (recall: a histogram only requires a single variable, e.g. `aes(x = carat)`)
- Save, stage, commit, pull, then push your changes back to your repo in RStudio **using the GUI interface (buttons)**
- Create a boxplot (see: `geom_box`) of diamond clarity (on the x-axis) and price (y-axis). 
- **In the command line**, stage, commit, & push changes


## End



