---
title: "Basic computation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic computation

Sometimes we use R as a basic calculator. Perform operations using expected symbols: 

- `*`: multiplication
- `/`: division
- `+/-`: addition/subtraction
- `^`: raise to this power
- `sqrt(x)`: take the square root of $x$
- `log(x)`: take the natural log of $x$
- `exp(x)`: natural exponential base (i.e. $e^x$)
- `<-`: assignment operator (to store outcomes, e.g. `cost <- price*units`)

### Example 1: Calculations with stored objects

The raw price of a wine glass is \$10.00. A winery is buying 350 glasses, which earns them a 15% discount for a bulk order. If there is an 8% sales tax on the discounted price, and a flat shipping rate of \$48.50, what is the total price the winery will pay? 

```{r}
# Create and store the variable values: 

cost_per_unit <- 10 # Cost per unit in dollars
units <- 350 # Total number of units ordered
discount <- 0.15 # 15% discount for buying > 300 units
tax_rate <-  0.08 # 8% sales tax 
shipping_cost <- 48.50 # Flat rate for domestic orders
```

**Critical thinking:** Why in some scenarios is it good practice to store values as objects, instead of type the value directly into an equation? 

OK, now that all of the values are assigned, let's find the discounted bulk price (there are a lot of different ways to calculate this): 

```{r}
discounted_price <- (1 - discount) * cost_per_unit * units
```

Then find the total tax on that discounted price:

```{r}
order_tax <- tax_rate*discounted_price
```

Now we have everything we need to find the total price!

```{r}
total_price <- discounted_price + order_tax + shipping_cost

total_price # Returns the total price
```

### Example 2: Exponents

A model for jaguar shark mass ($M$, kg) of the elusive jaguar shark is given by the following from Team Zissou, where $L$ is the shark length in meters:

$$M=0.4L^{2.6}$$

Using the model, what is the expected mass of jaguar sharks that is 4.8 meters long? (...yeah function examples below to make this more broadly useful & efficient, but here we'll just plug in the value.)

```{r}
# Store the shark length:
shark_length <- 4.8

# Calculate the shark mass using the model:
shark_mass <- 0.4 * (shark_length ^ 2.6)

# Return the expected shark mass
shark_mass
```

### Example 3: Evaluate a function 

Given the restricted function: $$P(t, S, \rho) = 0.4t\sqrt{4S+8.1}-e^{0.03t}+\ln{(6.2\rho - 3.9)}$$

What is $P(3.1, 2.5, 6.4)$?

*Note: couldn't you just write a function here, then plug in whatever you want for $t$, $S$, and $\rho$? Yes! You'll be writing a lot of functions in EDS 221 and beyond, but here's an example of this calculation done without a function.*

Let's start by assigning the variable values:

```{r}
t_val <- 3.1
s_val <- 2.5
rho_val <- 6.4
```

**Question:** Why might I have decided to use these object names, instead of something like $t$ and $s$?

Then write the equation to calculate $P$:

```{r}
p_val <- 0.4 * t_val * sqrt(4 * s_val + 8.1) - exp(0.3 * t_val) + log(6.2 * rho_val - 3.9)

# Return the calculated value:
p_val
```
**Question:** How is this function restricted? What happens if the $S$ value is updated to -10? If the $\rho$ value is 0.1? What are the possible *domains* for $S$ and $\rho$?
