---
title: "Summary statistics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
library(skimr)
```

In addition to exploratory data visualization, we often want to find basic summary statistics. Below are several tools & functions we use in EDS 212 to find them. 

### `skimr::skim()`

The `skimr::skim()` function returns "a broad overview of a data frame. It handles data of all types, dispatching a different set of summary functions based on the types of columns in the data frame."

```{r}
skimr::skim(penguins)
```

It also works on grouped data: 

```{r}
penguins %>% 
  dplyr::group_by(species) %>% 
  skimr::skim()
```

### `summary()`

As an alternative to `skimr::skim()`, the base `summary()` function provides basic information about variables. The type of summary information returned depends on the variable class for each column. 

```{r}
summary(penguins)
```

### `dplyr::group_by() %>% dplyr::summarize()`

The `group_by() %>% summarize()` combo is a powerful one to find summary statistics or outputs within groups, based on functions you specify. 

For example, first let's find the mean and standard deviation of flipper length, grouped by penguin species (`na.rm = TRUE` omits missing values): 

```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))
```
We can also group by more than one variable. For example, let's find the median bill length and sample size using `n()`, grouped by penguin species and year: 

```{r}
penguins %>% 
  group_by(species, year) %>% 
  summarize(median_bill_length = median(bill_length_mm, na.rm = TRUE),
            sample_size = n())
  
```

**Note:** If you are *only* finding counts of observations, you can use `dplyr::count()`, which bundles `group_by()`, `summarize()` and `n()` into one nice function. A comparison is below, to find the counts of penguins by species and island: 

```{r}
# Using group_by %>% summarize(n()):
penguins %>% 
  group_by(species, island) %>% 
  summarize(n())

# Using count():
penguins %>% 
  count(species, island)
```

