---
title: "Day 2: Derivatives"
description: |
  Definition of the derivative, basic rules, applications, interpretation, examples in EDS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 2 Part 1

### Morning lecture: 

**[Lecture 2A Slides - PDF](slides/EDS_212_day-2-part-1-slides.pdf)**

- Definition of the derivative
- Derivatives the long way
- Basic derivative rules

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-2-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Morning interactive session

**[Interactive Session 2A handout](comp_session_keys/comp_session_2a.html)**

- Derivatives by hand 
- Derivatives in R with `D()`
- Meet Jupyter Notebooks
- Hello Python

### Morning flex session: 

- git & GitHub continued

## Day 2 Part 2

### Afternoon lecture:

**[Lecture 2B Slides - PDF]()**

- Higher order derivatives
- Partial derivatives
- Applications in environmental data science

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-2-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Afternoon interactive session: 

**[Interactive Session 2B handout](comp_session_keys/comp_session_2b.html)**

- Derivatives in R continued
- Drawing higher order derivatives
- Partial derivatives in R

### Afternoon flex session: 

- Project oriented workflow
- Reproducible pieces sans code

## Day 2 Tasks and Activities

[Day 2 Tasks - Handout]()

## Efficiency tips

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

- `Cmd + A` (Mac) or `Ctrl + A`: Select all in current document
- `Cmd + comma` (Mac): Open preferences for current app

 
