---
title: "Day 1: Basic math review"
description: |
  Algebra review, unit conversions, interpreting graphs, univariate and multivariate functions, exponents, the natural exponential, and logarithms
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 1 Part 1

### Lecture topics

- Algebra basics refresher
- Units and unit conversions
- Functions: structure, language, evaluation
- Exponents

### Interactive session

- Meet R, RStudio, and R Markdown
- R as a calculator
- Storing objects
- Our first function


```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "https://allisonhorst.github.io/EDS_212_essential-math/slides/EDS_212_day-1-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

## Day 1 Part 2

### Lecture topics

- Natural exponential (& exponential functions)
- Logarithms
- Thinking about graphs
- Finding average rate of change (slope)

### Computational skills

- R Projects
- R Markdown customization
- Creating vectors and sequences
- Our first `ggplot2` graph
- Functions continued

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "https://allisonhorst.github.io/EDS_212_essential-math/slides/EDS_212_day-1-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```




