---
title: "Day 1: Basic math review"
description: |
  Algebra review, unit conversions, interpreting graphs, univariate and multivariate functions, exponents, the natural exponential, and logarithms
---

<!-- Note to self: Run the xaringan_to_pdf.R code to get function, then ran `xaringan_to_pdf("slides/EDS_212_day-1-part-1-slides.html")` to produce the PDF for these slides. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 1 Part 1 

### Lecture

**[Lecture 1.1 Slides - PDF](slides/EDS_212_day-1-part-1-slides.pdf)**

- Algebra basics refresher
- Units and unit conversions
- Functions: structure, language, evaluation
- Exponents

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-1-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Interactive session 

**[Interactive Session 1A handout](comp_session_keys/comp_session_1a.html)**

- Meet R, RStudio, and R Markdown
- Basic math operations and functions in R
- Storing objects
- Our first function

### Flex Session: 

- Meet git + GitHub

## Day 1 Part 2

### Lecture

**[Lecture 1.2 Slides - PDF](slides/EDS_212_day-1-part-2-slides.pdf)**

- Natural exponential (& exponential functions)
- Logarithms
- Thinking about graphs
- Finding average rate of change (slope)

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-1-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Interactive session

**[Interactive Session 1B handout](comp_session_keys/comp_session_1b.html)**

- R Projects
- R Markdown customization
- Creating vectors and sequences
- Our first `ggplot2` graph
- Functions continued

### Flex Session: 

- Interactive activity: expanding the Venn diagram of skills for data science

## Day 1 Tasks and Activities

### Refresher tasks

### Challenge problems

### Creative thinking problems

## Efficiency tips

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

- `Cmd + tab` (Mac) or `Ctrl + tab` (PC) to toggle through open applications
- `Ctrl + tab` (both) to toggle through tabs in browser (or open tabs in RStudio!)

