---
title: "Day 4: Linear algebra and summary statistics"
description: |
  Interpreting, writing, solving (basic) and using differential equations in EDS. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Day 4 Part 1:

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "https://allisonhorst.github.io/EDS_212_essential-math/slides/EDS_212_day-4-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

#### Flex session

- Fancy notebooks: R Markdown tips & tricks

## Day 4 Part 2: Summary statistics and data exploration

- Central tendency
- Spread & distributions
- Exploratory visualizations

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "https://allisonhorst.github.io/EDS_212_essential-math/slides/EDS_212_day-4-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

#### Flex session

- Watch party: JD Long's 2019 [RStudio::conf talk *Empathy in action*](https://www.youtube.com/watch?v=CjNOSrfQiAY)


## Day 4 Tasks and Activities

### Refresher problems

### Challenge problems

### Creative thinking problems

## Efficiency tips

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

