---
title: "Day 2: Derivatives"
description: |
  Definition of the derivative, basic rules, applications, interpretation, examples in EDS
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 2 Part 1

### Lecture: 

**[Lecture 2A Slides - PDF](slides/EDS_212_day-2-part-1-slides.pdf)**

- Definition of the derivative
- Derivatives the long way
- Basic derivative rules

### Interactive session

- Derivatives by hand 
- Meet Jupyter Notebooks
- Derivatives in R and Python
- Functions in Python

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-2-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

## Day 2 Part 2

### Lecture:

**[Slides - PDF]()**

- Higher order derivatives
- Partial derivatives
- Applications in environmental data science

### Interactive session: 

- Higher order derivatives as art
- Partial derivatives in R and Python
- Fun with partials & interpreting them


```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-2-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```
