---
title: "Day 5: Basic probability theory and Boolean algebra"
description: |
  Review of linear algebra basics, introduction to boolean algebra, and thinking like a computer
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 5 Part 1:

### Morning lecture

- Basic probability theory
- Finding probabilities from theoretical distributions
- Hypothesis tests - building intuition (& skepticism)


```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-5-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Morning interactive session

- Basic probability theory - problems 
- Probabilities from the normal distribution

### Morning flex session

- Start creating your personal website in `distill`

## Day 5 Part 2:

### Afternoon lecture

- Boolean algebra
- Notation & language
- Asking (`is.*`) versus telling (`as.*`) in code

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-5-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Afternoon interactive session

- Logic problems
- Common Boolean operators in R & Python

### Afternoon flex session

- Developing your personal `distill` website - continued

## Day 5 Tasks and Activities

[Day 5 Tasks - Handout]()

## Efficiency tips

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

- `Cmd + O`: Open Finder window
- `Cmd + Spacebar`: Open Spotlight Search

