---
title: "Day 3: Differential equations, intro to linear algebra"
description: |
  Differential equations, using them, integration concepts & basics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NOTE: Wednesday 10:00am late start

## Day 3 Part 1:

### Lecture: 

**[Lecture 3.1 Slides - PDF]()**

- Intro to differential equations 
- Differential equations in environmental data science
- Numerical solutions

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-3-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Interactive session (morning)

**[Interactive Session 3A handout](comp_session_keys/comp_session_3a.html)**

- Numerical solutions to differential equations
- Approximating solutions to the Lotke-Volterra equations in R

## Day 3 Part 2: Basic linear algebra for data science

**[Lecture 3.2 Slides - PDF]()**

- Language & shapes in linear algebra
- Basic operations
- How this translates to data structures & working with them

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-3-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

**Additional recommended reading:**

- [A gentle introduction to linear algebra](https://machinelearningmastery.com/gentle-introduction-linear-algebra/) by Jason Brownlee

### Interactive session (afternoon)

**[Interactive Session 3B handout](comp_session_keys/comp_session_3b.html)**

- Scalars, vectors, matrices, oh my!
- Basic inear algebra operations in R & Py 
    - Vector & matrix addition and subtraction
    - Scalar multiplication
    - Dot products
    - Matrix multiplication

## Day 3 Tasks and Activities

### Refresher problems

### Challenge problems

### Creative thinking problems

## Efficiency tips 

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

- `Cmd + C` (Mac) or `Ctrl + C` (PC): Copy
- `Cmd + X` (Mac) or `Ctrl + X` (PC): Cut
- `Cmd + V` (Mac) or `Ctrl + V` (PC): Paste
- `Cmd + Z` (Mac) or `Ctrl + Z`: Undo last
- `Cmd + Shift + Z` (Mac) or `Ctrl + Shift + Z` (PC): Redo last undo

