---
title: "Day 4: Linear algebra and summary statistics"
description: |
  Interpreting, writing, solving (basic) and using differential equations in EDS. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 4 Part 1:

### Morning lecture: 

**[Lecture 4A Slides - PDF]()**

- Linear algebra continued
- Matrix algebra
- Diagonals and transformation
- Linear algebra for systems of equations

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-4-part-1-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Morning interactive session

**[Interactive Session 4A handout]()**

- Basic linear algebra continued
- Summary statistics in R and Python
- Basic visual data exploration in `ggplot2`

### Morning flex session

- Fancy your notebooks: R Markdown tips & tricks

## Day 4 Part 2:

### Afternoon lecture: 

**[Lecture 4B Slides - PDF]()**

- Summary statistics and data exploration
- Central tendency
- Spread & distributions
- Exploratory data visualization

```{r, echo = FALSE}
xaringanExtra::embed_xaringan(
  url = "slides/EDS_212_day-4-part-2-slides.html",
  ratio = "16:9",
  border = "1px solid gainsboro"
)
```

### Afternoon interactive session

**[Interactive Session 4B handout]()**

- Summary statistics continued
- Data exploration continued

### Afternoon flex session

- Watch party: JD Long's 2019 [RStudio::conf talk *Empathy in action*](https://www.youtube.com/watch?v=CjNOSrfQiAY)

## Day 4 Tasks and Activities

[Day 4 Tasks - Handout]()

## Efficiency tips

*Inspired by Dr. Julia Lowndes' [Openscapes](openscapes.org) program.*

- `Cmd + F`: Open 'Find and replace' in current doc
- `Cmd + Shift + F`: Search in entire R project

