---
title: 'EDS 212 Computational Session 2B'
subtitle: 'Higher order derivatives, partial derivatives, plotting functions in ggplot'
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r, echo = TRUE}
library(tidyverse)
```

## Derivatives by hand

1. Find all partial derivatives of the following: 

$$f(a, z, t)=3at^2-4t+5.1az-z^3$$

2. Algal biomass ($B$, grams) in a system is modeled by the following, where $h$ is time (hours) and $T$ is temperature (Celsius). 

$$B(h, T)=0.6Th+2.1h^2$$
For water at 28 degrees, find an expression for how biomass is changing with respect to time ($\frac{\partial B}{\partial h}$). 

3. Find the 3^rd^ derivative of the following: 

$$G(t)=2.2 + 3.1t-5.6t^4 + \frac{4.3}{t^3}$$
## Plotting a function in `ggplot2`

```{r}
eq = function(x){3*x^2 + 4} # Define function

# Plot it over a specific range using the `geom_function` mark: 
ggplot(data.frame(x = c(1, 100)), aes(x = x)) + 
  geom_function(fun = eq)

```

```{r}
eq_2 = function(x) {2.4 - 5*x^3 + 2.1*x^2} # Define function

ggplot(data.frame(x = c(-50,50)), aes(x = x)) + 
  geom_function(fun = eq_2)
```

## Plotting derivatives

Let's look at the function: $C(t)=t^3$

Let's plot the function along with the first derivative: 

```{r}
# Store the function C(t)
ct <- function(t) {t ^ 3}

# Plot it:
ggplot(data.frame(t=c(-5, 5)), aes(x=t)) + 
  geom_function(fun=ct)

# Find the derivative:
dc_dt <- D(expression(t^3), 't')
dc_dt

# Store the derivative as a function:
dc_dt_fun <- function(t) {3 * t ^2}

# Then plot them together! 
ggplot(data.frame(t = c(-5, 5)), aes(x = t)) + 
  geom_function(fun = ct, color = "red") +
  geom_function(fun = dc_dt_fun, color = "blue")
```

## Partial derivatives in R

Specify the variable you want to take the derivative with respect to. 

$$f(x,y,z)=2xy-3x^2z^3$$

Find all partials, $\frac{\partial f}{\partial x}$, $\frac{\partial f}{\partial y}$ and $\frac{\partial f}{\partial z}$

```{r}
f_xyz <- expression(2*x*y - 3*(x^2)*(z^3))

# Partial with respect to x: 
df_dx <- D(f_xyz, 'x')
df_dx

# Partial with respect to y: 
df_dy <- D(f_xyz, 'y')
df_dy

# Partial with respect to z: 
df_dz <- D(f_xyz, 'z')
df_dz
```

## One more example:

Given the function: $$P(q)=2cos(3q+0.4)-5.6e^{1.4q}$$ 

Find the instantaneous slope at $q = 0.8$.

```{r}
# First, create the expression: 
pq <- expression(2 * cos(3 * q + 0.4) - 5.6 * exp(1.4 * q))

# Find the first derivative with respect to q:
dp_dq <- D(pq, 'q')

# Return the first derivative:
dp_dq

# Define value of q
q <- 0.8

# Evaluate dp_dq at that value: 
eval(dp_dq)
```

## End {-}
