---
title: 'EDS 212 Computational Session 3A'
subtitle: 'Differential equations and solving them numerically in R: Lotke-Volterra'
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
  
---

In this session, we'll see how to use the `deSolve::ode()` function to numerically estimate the solution to the Lotke-Volterra equations for predator-prey populations.

First, we'll learn a bit more about how numerical methods for approximating solutions to differential equations work. 

**Credit:** This lesson is closely based on the article (Numerically solving differential equations with R)[https://rstudio-pubs-static.s3.amazonaws.com/32888_197d1a1896534397b67fb04e0d4899ae.html]

```{r}
library(tidyverse)
library(deSolve)
library(kableExtra)
```

## Numerical approximation by hand

We'll just do this once to reinforce how numerical integration generally works. Given the differential equation:

$$\frac{dP}{dt}=0.4P(0.5 - P)$$

and the initial condition $P(0)=0.2$, approximate the solution $P(t)$ at t = 1, 2, 3, 4, 5. 

First, what is $t$, $P$, and $\frac{dP}{dt}$ at $t = 0$? Here we have the initial condition, and can calculate the slope: 

```{r}
t_0 <- data.frame(t = 0, P = 0.2, 'dfdx' = 0.024)

kable(t_0) %>% 
  kable_styling(full_width = FALSE)
```

Now, we approximate the value of $P(t)$ and $\frac{dP}{dt}$ at the next step ($t = 1$) *as if the slope is continuing at 0.024*. Which means the approximate value of $P(1) = 0.2 + 0.024*(1) = 0.44$. At that value of $P$, the slope is approximated by $\frac{dP}{dt}=0.4*0.44(0.5 - 0.44) = 0.01056$.

So now, our table is expand to: 

```{r}
t_1 <- data.frame(t = c(0,1), P = c(0.2, 0.44), 'dfdx' = c(0.024, 0.01056))

kable(t_1) %>% 
  kable_styling(full_width = FALSE)
```

Then we continue. What do we approximate is the value of P(t) at t = 2? Using the new slope of 0.01056, the approximate value of $P(2) = 0.44 + 0.01056*1 = 0.45056$. Using that value, we approximate the slope at $t=2$ by $\frac{dP}{dt}=0.4*0.45056(0.5 - 0.45056) = 0.0089$. 

Our table is then updated to: 

```{r}
t_2 <- data.frame(t = c(0,1,2), P = c(0.2, 0.44, 0.45056), 'dfdx' = c(0.024, 0.01056, 0.0089))

kable(t_2) %>% 
  kable_styling(full_width = FALSE)
```
And we could continue doing that for many incremental values of $t$ to get an approximate solution for what the function $P(t)$ looks like. But that would be *very* tedious. Up next - computer help! 

### Summary: numerical approximation
There are *many* different approaches to numerical approximation of solutions for differential equations. This example is meant to give you an idea of how the iterative approximation works generally.

Now, let's see how much more quickly we can do this with some computer power! 

## Numerical approximation using `deSolve::ode()`

### The Lotke-Volterra equations

As described in the lecture, the Lotke-Volterra models have been used to describe predator-prey populations. 

#### Prey equation: 

$$\frac{dx}{dt}=\alpha x-\beta xy$$

From Wikipedia: "The prey are assumed to have an unlimited food supply and to reproduce exponentially, unless subject to predation; this exponential growth is represented in the equation above by the term $\alpha x$. The rate of predation upon the prey is assumed to be proportional to the rate at which the predators and the prey meet, this is represented above by $\beta xy$. If either x or y is zero, then there can be no predation."

#### Predator equation: 

$$\frac{dy}{dt}=\delta xy - \gamma y$$

From Wikipedia: "In this equation, $\delta xy$ represents the growth of the predator population. (Note the similarity to the predation rate; however, a different constant is used, as the rate at which the predator population grows is not necessarily equal to the rate at which it consumes the prey). The term $\gamma y$ represents the loss rate of the predators due to either natural death or emigration, it leads to an exponential decay in the absence of prey. 

Where: 

- $x$ is prey population (e.g. rabbits)
- $y$ is predator population (e.g. wolves)
- $\alpha, \beta, \gamma, \delta$ are positive parameters

To find an approximate solution in R, we will need four things: 
- Parameter values
- A sequence of times over which we'll approximate predator & prey populations
- An initial condition (initial populations of predator & prey at t = 0)
- The differential equations that need to be solved

Solving the Lotke-Volterra equation: 

```{r}
# Create a sequence of times (days): 
time <- seq(0, 25, by = 0.05)

# Set some parameter values (these can change - keep it in mind):
parameters <- c(alpha = .75, beta = 0.8, delta = 0.5, gamma = 1)

# Set the initial condition (prey and predator populations at t = 0).
# Recall: x = prey, y = predator
init_cond <- c(x = 5, y = 2)

# Prepare the series of differential equations as a function: 
lk_equations <- function(time, init_cond, parameters) {
  with(as.list(c(init_cond, parameters)), {
    dxdt = alpha * x - beta * x * y
    dydt = delta * x * y - gamma * y
    return(list(c(dxdt, dydt)))
  })
}

# Find the approximate the solution using `deSolve::ode()`:
approx_lk <- ode(y = init_cond, times = time, func = lk_equations, parms = parameters)

# Check the class: 
class(approx_lk)

# We really want this to be a data frame, and we want both prey (x) and predator (y) to be in the same column -- we'll learn why in EDS 221 (tidy data)
approx_lk_df <- data.frame(approx_lk) %>% 
  pivot_longer(cols = c(x,y), names_to = "species", values_to = "population")

# Plot it! 
ggplot(data = approx_lk_df, aes(x = time, y = population)) +
  geom_line(aes(color = species))
```

## Updating parameter values

What happens as you change the different parameters (and re-run the entire code chunk)? How does that align with what you see in the graph? Some things to keep in mind: 

- $\alpha$ is a growth rate for prey
- $\gamma$ is a mortality rate for predator

## End
