---
title: 'EDS 212 Computational Session 5B'
subtitle: 'Logical and comparison operators in R'
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
---

## Setup

- Close, then reopen your project from the earlier session
- Pull to get any changes since you last worked on it
- Create a new R SCRIPT 
- Save your script as `intro-logicals.R`

## Comparison operators

Comparison (or relational) operators compare values and return `TRUE` or `FALSE` depending on what you're comparing. 

### Less than or greater than

```{r}
2 < 4
```

```{r}
5 < 2
```

```{r}
"A" < "D"
```

```{r}
"Z" < "Q"
```

```{r}
"cat" < "dog"
```

```{r}
"yeti" < "armadillo"
```

So we can see that values can be considered numerically or alphabetically. 

This is also true for vectors - each element in a vector will be compared to the corresponding value in the other vector (i.e., the first element in `a` gets compared to the first element in `b`). Let's make a couple vectors and try it out: 

```{r}
dogs <- c(1, 10, 4, 5, 18)
cats <- c(0, 12, 10, 5, 2)

dogs < cats
```

Same for vectors of strings: 
```{r}
dog_names <- c("Teddy", "Khora", "Banjo", "Waffle")
cat_names <- c("Henrietta", "Panda", "Merlin", "Runt")

cat_names < dog_names
```


### Less than / greater than **or equal to**

```{r}
2 <= 5
```

```{r}
10 >= 10
```

Test alphabetically: 

```{r}
"yeti" <= "yeah"
```

Let's try it with vectors: 

```{r}
burritos <- c(-4, 10, 1, 0.6, 9)
tacos <- c(5, 3, 8, -5, 6)

tacos <= burritos
```

### Exact matches with `==`

Use `==` to ask "do these things match?" 

```{r}
5 == 5
```

```{r}
5 == 4
```

```{r}
"cat" == "dog"
```

```{r}
beach <- c(0, 2, 1, 10, 13)
mesa <- c(1, 2, 3, 10, 13)

beach == mesa
```

Let's try something interesting: 
```{r}
5 == TRUE

5 == FALSE

1 == TRUE

0 == TRUE

0 == FALSE
```

0 recognized as 'FALSE' and 1 recognized as 'TRUE'. 

### Does NOT match with `!=`

Use `!=` to ask: does this NOT match that?

```{r}
5 != 6
```

```{r}
"cat" != "dog"
```

## Logical operators

- `&`: logical "AND"
- `|`: logical "OR"
- `!`: logical "NOT"

### Logical "AND"

```{r}
5 < 10 & "cat" < "dog"
```

```{r}
9 > 4 & 2 < 3
```

An example with vectors: 

```{r}
bananas <- c(10, 21, 12, 15, 22)
apples <- c(9, 18, 16, 20, 30)

apples < bananas & bananas > 20
```


### Logical "OR"

```{r}
4 < 10 | 100 < 2
```

```{r}
5 > 6 | "dog" < "cat" | 10 < 50
```

### Logical "NOT"

```{r}
!"dog" == "cat"

!6 == 5

```

