---
title: 'EDS 212 Computational Session 4B'
subtitle: 'Exploring data in R and Python'
output: 
  html_document:
    css: comp_session_theme.css
    toc: true
    toc_float: true
---


```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(reticulate)
library(GGally)
library(skimr)
library(palmerpenguins)
```


## Set up

- Create a new repository on GitHub called `eds212-comp-4b`, with a ReadMe
- Clone and create a new version-controlled R Project
- Add a new R Markdown document, save as `r-exploring`
- Open Anaconda Navigator, launch Jupyter Notebook, navigate to the Project folder you just created, & make a new Jupyter Notebook there w/Python3 (when you save your Jupyter NB, make sure it's saving *in your R project*)
- Rename your Jupyter Notebook `py-exploring`
- Check your files pane in RStudio to ensure that your .ipynb is saved in the right place

## Exploring data in R:

In your RMarkdown document, attach the following packages in the setup chunk (you'll need to install the first two):

- `GGally`
- `skimr`
- `palmerpenguins`

Rapid-fire low-level exploration of data:
```{r}
# Always look at it
# View(penguins)

# Check the column names
names(penguins) # See df.columns in pandas

# Check the dimensions 
dim(penguins) # See df.shape in pandas

# Get a summary
summary(penguins) # See df.describe() in pandas

# Print the first 6 lines
head(penguins) # See df.head() in pandas

# Print the last 6 lines 
tail(penguins) # See df.tail() in pandas

# Make a pairplot
GGally::ggpairs(penguins)

# Make a histogram of penguin flipper lengths
ggplot(data = penguins, aes(x = flipper_length_mm)) +
  geom_histogram()
```

## Let's try the same stuff, but in Python! 

```{python}
# Import Python packages
import seaborn as sns
import pandas as pd
import numpy as np
```

```{python}
# Load the penguins dataset from the seaborn package
penguins = sns.load_dataset('penguins')
```

```{python}
penguins.columns # See names(penguins) in R
 
penguins.shape # See dim(penguins) in R

penguins.head() # See head(penguins) in R

penguins.tail() # See tail(penguins) in R

penguins.describe() # See summary(penguins) in R

# Make a pairs plot with seaborn pairplot
sns.pairplot(penguins) # See GGally::ggpairs() in R

# Make a histogram of flipper lengths with sns.histplot:
sns.histplot(data=penguins, x="flipper_length_mm") # See geom_histogram() in R
```

One more thing: vectors in Python 

```{python}
vec_a = np.array([1,2,3])
vec_b = np.array([10,11,12])

vec_a + vec_b
vec_b - vec_a
vec_a * vec_b
6 * vec_a
```

